Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-05-25 SIMPaper/sbpaper.Rmd
cleaned, licensed version of cbnn

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "In survival analysis, covariate interactions are difficult to model using regression-based methods without prior knowledge. Neural network-based survival models have been developed for data-driven interaction modeling, however, these approaches lack implementations for time-varying interactions or a flexible baseline hazard. To address this, we extend the case-base sampling framework and propose Case-Base Neural Networks (CBNN), which introduce time directly into the model, using commonly available neural network components for a simple implementation. First, we case-base sample the survival data, converting the survival time of each participant into discrete moments in time. This allows us to estimate the probability an event occurs at a given moment using a neural network to get the hazard. We compare CBNN to survival methods based on regression and neural networks in two simulations and two real data applications with a fixed neural network architecture. We report metrics for each model over follow-up time using both a proper (Index of Prediction Accuracy) and non-proper (inverse probability censoring weights-adjusted concordance index) scoring rule. A simple simulation shows CBNN outperforming the other neural network methods. The complex simulation highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, outperforming all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. CBNN has shown that it is more consistent in prediction performance across time in the simulations and real data applications compared to the competing neural network approaches."

@@

In survival analysis, covariate interactions are difficult to model using regression-based methods without prior knowledge. Neural network-based survival models have been developed for data-driven interaction modeling, however, these approaches can not model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) which extends the case-base sampling framework which introduce time directly into the model. We case-base sample the survival data, converting the survival time of each participant into discrete moments in time. This allows us to use a feed-forward neural network that estimate the probability of an event occurring at a given moment, which can be used to estimate the hazard function. We then compare CBNN to survival methods based on regression and neural networks in two simulations and two real data applications with a fixed neural network architecture. We report metrics for each model over follow-up time using both proper (Index of Prediction Accuracy) and non-proper (inverse probability censoring weights-adjusted concordance index) scoring rules. A simple simulation with an exponential hazard model shows CBNN consistently outperforming the other neural network methods. The complex simulation highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, outperforming all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time than the competing neural network approaches.

# Introduction

what is aft, dont like assuming distribution, smooth. Less used than cox. Proportional hazards assumptions hard to maintain in long study, all regression require prior knowledge.

The widespread use of the Cox proportional hazards model has led to a focus on survival models based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009].
In clinical settings, Cox proportional hazards models are frequently prioritized over smooth-in-time accelerated failure time (AFT) models [@hanley2009], which can estimate absolute risks by modeling the hazard directly through user specification of a baseline hazard distribution. Using splines as the baseline hazard is an alternative to assuming a specific distribution [@royston2002flexible]. The risk associated with each covariate may not be constant over time, making it difficult to maintain the proportional hazards assumption. Previous studies related to breast cancer have discovered time-varying interactions with covariates of interest, like tumor size [@coradini2000time]. A limitation of regression-based models is the requirement of prior knowledge; the analyst must know potential time-varying interactions and explicitly evaluate them.

Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through user specification of a baseline hazard distribution [@kleinbaum2012survival]. The assumption of a specific distribution tends to be a point of contention. A method has been developped using splines, providing flexibility to the baseline hazard  [@royston2002flexible]. AFT models are underutilized compared to Cox proportional hazards, resulting in a focus on survival models based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. However, The risk associated with each covariate may not be constant over time, making it difficult to maintain the proportional hazards assumption. Previous studies related to breast cancer have discovered time-varying interactions with covariates of interest, like tumor size [@coradini2000time]. A limitation of all regression-based models is the requirement of prior knowledge; the analyst must know potential time-varying interactions and explicitly evaluate them.

Neural networks provide a data-driven approach to approximating interaction terms. DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood as a custom loss function [@katzman2018DeepSurv]. This results in a stepwise absolute risk curve that cannot accommodate time-varying interactions. Compared to Cox regression, DeepSurv has shown an improvement in performance on the Study to Understand Prognoses Preferences and Risk Treatments (SUPPORT) dataset [@knaus1995SUPPORT]. Farraggi et al. suggested a modification to the loss function required for a Cox neural network that can handle non-proportional hazards [@faraggi1995neural]. Wang et al. took the concept of extreme learning machines and applied them in a survival context, removing the need for tuning the number of layers and nodes [@wang2019extreme].

DeepHit requires that each survival time of interest be specified in the model and directly estimates survival curves, rather than deriving a hazard function [@lee2018DeepHit]. It assumes an inverse Gaussian distribution as the baseline hazard and its concordance index (c-index) score outperforms DeepSurv [@katzman2018DeepSurv] on the Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) dataset [@curtis2012genomic].

Deep Survival Machines (DSM) provides a flexible, parametric survival model using neural networks with a mixture of distributions as the baseline hazard [@dsmPaper]. On both the SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit [@dsmPaper]. However, like DeepHit and DeepSurv, DSM cannot model time-varying interactions.

These methods need custom loss and activation functions that are difficult to implement in practice. We note the complexity of these alternative neural network approaches; DeepSurv, DeepHit, and DSM all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper], while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. Following the case-base framework, we can use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires user specification of time-varying interactions. CBNN can model both without user specification; only a sufficiently deep network is required.

In Section \ref{methods}, we describe how case-base sampling and neural networks are combined both conceptually and algebraically, along with our hyperparameter choices and software implementation. In Section \ref{sims}, we describe our metrics and compare the performance of CBNN, DeepSurv, DeepHit, DSM, Cox regression, and case-base using logistic regression on simulated data. Section \ref{casestudies} describes the real-data analysis, while in Section \ref{discussion}, we explore the implications of our results and contextualize them within neural network survival analysis in a single event setting.

# Case-base neural network methodology, metrics, and software {#methods}

@@

For each person-moment sampled, let $X_i$ be the corresponding covariate profile $\left(x_{i1},x_{i2},...,x_{ip} \right)$, $T_i$ be the time of the person-moment, and $Y_i$ be the indicator variable for whether the event of interest occurred at time $T_i$. We estimate the hazard function $h(t \mid X_i)$ using the sampled person-moments. Recall that $h(t \mid X_i)$ is the instantaneous potential of experiencing the event at time $t$ for a given set of covariates $X_i$, assuming $T_i \geq t$.

Now, let $b$ be the (user-defined) size of the *base series*, and let $B$ be the sum of all follow-up times for the individuals in the study. If we sample the *base series* uniformly, then the hazard function of the sampling process is equal to $b/B$. Therefore, we have the following equality \footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}: \begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} 
The odds of a person-moment being a part of the *case series* is the ratio of the hazard $h(T_i \mid X_i)$ and the uniform rate $b/B$. Using \eqref{eqn:main}, we can see how the log-hazard function can be estimated from the log-odds arising from case-base sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

We may estimate the hazard function if we adjust for the bias introduced when sampling a fraction of the study base $B$ by adding the offset term $\log\left(\frac{b}{B} \right)$ (\eqref{eqn:offset}). Next, we propose using neural networks to model the odds.

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By allowing a neural network to approximate a higher-order polynomial of time, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of nodes, and layers control the flexibility of the model. We provide a detailed description of the choices we made in the next sub-section.

@@

The following derivation shows how our probability estimate is converted to odds:
\begin{align*}
 \log\left( h(t \mid X_i) \right) &= \log\left(\frac{\sigmoid \left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\sigmoid \left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function [@gulli2017]:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*}
where $\hat{f}_{\theta}(x_{i}, t_{i})$ is our estimate for a given covariate profile and time, $y_{i}$ is our target value specifying whether an event occurred, and $N$ represents the number of individuals in our training set.

Backpropagation is used to optimize the parameters in the model with an appropriate minimization algorithm (e.g. Adam, RMSPropagation, stochastic gradient descent)[@gulli2017]. For our analysis, we use Adam. Note that the size of the *case series* is fixed as the number of events, but we can make the *base series* as large as we want. A ratio of 100:1 *base series* to *case series* is sufficient [@hanley2009]. We pass our feed-forward neural network through a sigmoid activation function (Figure \ref{fig:NNarch}). Finally, we can convert this model output to a hazard. When using our model for predictions, we manually set the offset term to 0 in the new data, as we account for the bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the risk function ($F$) at time $t$ for a covariate profile $X$, viz.\begin{align}\label{eqn:ci2}
F\left(t|X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} 
We use Riemann's Sum [@hughes2020calculus] to approximate the integral in \eqref{eqn:ci2}.

@@

Neural networks provide flexibility when defining the architecture and optimization parameters of our models. These hyperparameter choices can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a validation split of $20\%$, a testing split of $10%$, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs, and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test). We select the best weights after training based on validation loss [@gulli2017].

@@

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DSM model using **DeepSurvivalMachines** [@dsmPaper] and the DeepHit model using **pycox** [@lee2018DeepHit]. We made the components of CBNN using the **keras** [@keras] package and the **casebase** package for the sampling step. The **simsurv** package [@simsurv] is used to simulate our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of c-ipcw from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Brier score and IPA metrics. We changed the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

@@

In this section, we use simulated data to evaluate the performance of CBNN and compare our approach with existing methods based on regression (Cox, CBLR) and neural network (DeepHit, DeepSurv, DSM) methods. We specify a linear combination of each covariate as the linear predictor in regression-based approaches (Cox, CBLR), which contrasts with neural network approaches that allow for non-linear interactions. We simulate data under a simple exponential model, and a complex baseline hazard with time-varying interactions, each with 10% random censoring. For both settings, we simulate three covariates:

@@

The DeepHit-specific hyperparameter alpha is set to $0.5$ (equal weight between its negative log-likelihood and ranking losses [@lee2018DeepHit]). We change the **DeepSurvivalMachines** [@dsmPaper] package to include dropout and a minimum delta loss during the fitting process. For DSM, we define a mixture model of six Weibull distributions for the baseline hazard. All other hyperparameters are held constant across all neural network methods in both the simulation studies and real data applications.

Besides the methods mentioned above, we include the Optimal model in our comparisons using CBLR. That is, we include the exact functional form of the covariates in a CBLR model (referred to as Optimal for simplicity). We calculate $t$-based 95% confidence intervals using 100 replications of the simulated data For all analyses, we use 80% for training and 20% for the test set. 20% of the training set is kept for validation at each epoch. We predict risk functions $F$ using \eqref{eqn:ci2} for individuals in the test set, which are used to calculate our c-ipcw and IPA scores.

@@

We use two metrics to assess the performance of the different methods of interest: 1) index of prediction accuracy (IPA) [@kattan2018index] and 2) inverse probability censoring weights-adjusted concordance index (c-ipcw) [@uno2011], which we define below. Both these time-dependent metrics provide transparency as to when in follow-up time each model may perform better than the others.

@@

The IPA is a function of the Brier score ($BS(t)$) [@graf1999], which is defined as
\begin{align}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(\widehat{F}(X_{i},t)-1\right)^{2}\cdot I_{T_{i}\leq t,\delta_{i}=1}}{\widehat{G}(T_{i})} + \frac{\left(\widehat{F}(X_{i},t)\right)^{2}\cdot I_{T_{i}>t}}{\widehat{G}(t)}\right),
\end{align}
where $\delta_{i}=1$ shows individuals who have experienced the event, $N$ represents the number of samples in our dataset over which we calculate $BS(t)$, $\widehat{G}(t)=P[C>t]$ is a non-parametric estimate of the censoring distribution, $C$ is censoring time and $T_{i}$ is an individual’s survival or censoring time. The Brier score provides a score that accounts for the information loss because of censoring. There are three categories of individuals that may appear within the dataset once we fix our $t$ of interest. Individuals who experienced the event before $t$ are present in the first term of the equation. The second term of the equation includes individuals who experience the event or are censored after $t$. Those censored before $t$ are the third category of people. The  inverse probability censoring weights (IPCW) adjustment ($G(\cdot)$) is to account for these category three individuals whose information is missing. The IPA score as a function of time is given by
\begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where $BS_{model}(t)$ represents the Brier score over time $t$ for the model of interest, and $BS_{null}(t)$ represents the Brier score if we use an unadjusted Kaplan-Meier (KM) curve as the prediction for all observations [@kattan2018index]. Note that IPA has an upper bound of one, where positive values show an increase in performance over the null model and negative values show that the null model performs better. These scores demonstrate how performance changes over follow-up time.

A potential artifact of IPA is that the score is unstable at earlier and later survival times. This is because of a near equivalent Brier score between each model and the null model. At small values, a difference of $0.1$ creates a more significant fold change than at larger values. As the Brier score is potentially small at the start and at the end of their respective curves, we may see instability of the IPA score at those ends.

@@

The c-ipcw is a non-proper, rank-based metric that does not depend on the censoring times in the test data [@uno2011]. The c-ipcw is given by

>>>>

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "Neural network-based survival methods have been developed to model data-driven covariate interactions. While these methods have led to better predictive performance than regression-based approaches, they cannot model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling framework with flexible architectures. Our method naturally accounts for censoring, does not require additional hyperparameters and uses binary cross-entropy as the loss function. Using a novel sampling scheme and data augmentation, we incorporate time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a given moment and estimates the hazard function. We compare the performance of CBNN to survival methods based on regression and neural networks in two simulations and two real data applications. We report two time-dependent metrics for each model. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time and outperforms the competing neural network approaches. For a simple simulation with an exponential hazard model, CBNN consistently outperforms the other neural network methods. For a complex simulation, which highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, CBNN outperforms all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. This study highlights the benefit of combining case-base sampling with deep learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling of survival outcomes."

@@

# Introduction
Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through user-specified baseline hazard distribution [@kleinbaum2012survival]. Cox proportional hazards models are used more often than AFT models, causing analyses to be based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. The identification of an appropriate distribution for the baseline hazard in an AFT model may be difficult for common diseases that have many interacting risk factors or a Cox model whose pathogenesis may change with age, making it difficult to maintain the proportional hazards assumption. For example, previous studies of breast cancer incidence have discovered time-varying interactions with covariates of interest, including tumor size [@coradini2000time]. One approach to resolve this involves using the basis of splines on time to provide flexibility in the baseline hazard [@royston2002flexible]. However, regression-based models are limited in that they require prior knowledge about potential time-varying interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms. For example, DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood as a custom loss function [@katzman2018DeepSurv], resulting in a stepwise absolute risk curve that cannot accommodate time-varying interactions. Compared to Cox regression, DeepSurv shows better performance on the Study to Understand Prognoses Preferences and Risk Treatments (SUPPORT) dataset [@knaus1995SUPPORT]. To handle non-proportional hazards, a modification to the loss function was proposed [@faraggi1995neural]. To remove the need for tuning the number of layers and nodes, the concept of extreme learning machines has been applied to a Cox neural network model [@wang2019extreme].

As an alternative method that assumes a baseline hazard distribution, DeepHit specifies each survival time of interest in the model and directly estimates survival curves, rather than deriving a hazard function [@lee2018DeepHit]. It assumes an inverse Gaussian distribution as the baseline hazard and it outperforms DeepSurv on the Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) dataset [@curtis2012genomic].

Providing further flexibility, Deep Survival Machines (DSM) is a parametric survival model using neural networks with a mixture of distributions as the baseline hazard [@dsmPaper]. On both the SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit [@dsmPaper]. However, like DeepHit and DeepSurv, DSM cannot model time-varying interactions.

We note that these alternative neural network approaches all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper]. DeepHit introduces a new hyperparameter weighing its two loss functions (negative log-likelihood and ranking losses) while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull  [@lee2018DeepHit] [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. The current neural network models provide flexibility at the cost of opacity, while regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. As part of the case-base framework, we use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires explicit use of time-varying interactions. CBNN can model both without user specification; only a sufficiently deep network is required.

In Section \ref{methods}, we describe how case-base sampling and neural networks are combined both conceptually and algebraically, along with our hyperparameter choices and software implementation. Section \ref{sims}, describes our metrics and compares the performance of CBNN, DeepSurv, DeepHit, DSM, Cox regression and case-base using logistic regression (CBLR) on simulated data. Section \ref{casestudies} describes the real-data analysis, while Section \ref{discussion} explores the implications of our results and contextualizes them within neural network survival analysis in a single event setting.

# Case-base neural network methodology, metrics and software {#methods}

@@

For each person-moment sampled, let $X_i$ be the corresponding covariate profile $\left(x_{i1},x_{i2},...,x_{ip} \right)$, $T_i$ be the time of the person-moment and $Y_i$ be the indicator variable for whether the event of interest occurred at time $T_i$. We estimate the hazard function $h(t \mid X_i)$ using the sampled person-moments. Recall that $h(t \mid X_i)$ is the instantaneous potential of experiencing the event at time $t$ for a given set of covariates $X_i$, assuming $T_i \geq t$.

Now, let $b$ be the (user-defined) size of the *base series* and let $B$ be the sum of all follow-up times for the individuals in the study. If we sample the *base series* uniformly across the study base, then the hazard function of the sampling process is equal to $b/B$. Therefore, we have the following equality \footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}: 
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} 
The odds of a person-moment being a part of the *case series* is the ratio of the hazard $h(T_i \mid X_i)$ and the uniform rate $b/B$. Using \eqref{eqn:main}, we can see how the log-hazard function can be estimated from the log-odds arising from case-base sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjusting for the bias introduced when sampling a fraction of the study base $B$ by adding the offset term $\log\left(\frac{b}{B} \right)$ as in \eqref{eqn:offset}. Next, we propose using neural networks to model the odds.

@@

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By approximating a higher-order polynomial of time using a neural network, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of layers and nodes control the flexibility of the model. We provide a detailed description of the choices we made in the next sub-section.

@@

The following derivation shows how our probability estimate is converted to odds:
\begin{align*}
 \log\left( h(t \mid X) \right) &= \log\left(\frac{\sigmoid \left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\sigmoid \left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function [@gulli2017]:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*}
where $\hat{f}_{\theta}(x_{i}, t_{i})$ is our estimate for a given covariate profile and time, $y_{i}$ is our target value specifying whether an event occurred and $N$ represents the number of individuals in our training set.

Backpropagation with an appropriate minimization algorithm (e.g. Adam, RMSPropagation, stochastic gradient descent) is used to optimize the parameters in the model [@gulli2017]. For our analysis, we use Adam as implemented in Keras [@gulli2017]. Note that the size of the *case series* is fixed as the number of events, but we can make the *base series* as large as we want. A ratio of 100:1 *base series* to *case series* is sufficient [@hanley2009]. We pass our feed-forward neural network through a sigmoid activation function (Figure \ref{fig:NNarch}). Finally, we can convert this model output to a hazard. When using our model for predictions, we manually set the offset term to 0 in the new data, as we account for the bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the risk function ($F$) at time $t$ for a covariate profile $X$, viz.\begin{align}\label{eqn:ci2}
F\left(t\mid X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} 
We use a finite Riemann sum [@hughes2020calculus] to approximate the integral in \eqref{eqn:ci2}.

@@

Neural networks are flexible when defining the architecture and optimization parameters. These hyperparameter decisions can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a validation split of $20\%$, a testing split of $10%$, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test) for each method. We select the best weights after training based on validation loss [@gulli2017].

@@

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DeepHit model using **pycox** [@lee2018DeepHit] and the DSM model using **DeepSurvivalMachines** [@dsmPaper]. We made the components of CBNN using the **casebase** package for the sampling step and the **keras** [@keras] package for our neural network architecture. The **simsurv** package [@simsurv] is used for our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of $C_{IPCW}$??? from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Index of Prediction Accuracy (IPA metric). Both metrics are described in detail in the following section. We modify the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

@@

In this section, we use simulated data to evaluate the performance of CBNN and compare our approach with existing regression-based (Cox, CBLR) and neural network-based (DeepHit, DeepSurv, DSM) methods. We specify a linear combination of each covariate as the linear predictor in regression-based approaches (Cox, CBLR), which contrasts with neural network approaches that allow for non-linear interactions. We simulate data under a simple exponential model and a complex baseline hazard with time-varying interactions, each with 10% random censoring. For both settings, we simulate three covariates:

@@

The DeepHit-specific hyperparameter alpha is set to $0.5$ (equal weight between its negative log-likelihood and ranking losses [@lee2018DeepHit]). We modify the **DeepSurvivalMachines** [@dsmPaper] package to include dropout and a minimum delta loss during the fitting process. For DSM, we define a mixture model of six Weibull distributions for the baseline hazard. All other hyperparameters are held constant across all neural network methods in both the simulation studies and real data applications.

Besides the methods mentioned above, we include the Optimal model in our comparisons using CBLR. That is, we include the exact functional form of the covariates in a CBLR model (referred to as Optimal for simplicity). We calculate $t$-based 95% confidence intervals using 100 replications of the simulated data. For all analyses, we use 80% for training and 20% for the test set. 20% of the training set is kept for validation at each epoch. We predict risk functions $F$ using \eqref{eqn:ci2} for individuals in the test set, which are used to calculate our $C_{IPCW}$ and IPA scores.

@@

We use two metrics to assess the performance of the different methods of interest: 1) (IPA) [@kattan2018index] and 2) inverse probability censoring weights-adjusted concordance index ($C_{IPCW}$) [@uno2011], which we define below. Both these time-dependent metrics provide transparency as to when in follow-up time each model may perform better than the others.

@@

The IPA is a function of the Brier score ($BS(t)$) [@graf1999], which is defined as
\begin{align}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align}
where $\delta_{i}=1$ shows individuals who have experienced the event, $N$ represents the number of samples in our dataset over which we calculate $BS(t)$, $\widehat{G}(t)=P[c>t]$ is a non-parametric estimate of the censoring distribution, $c$ is censoring time and $T_{i}$ is an individual’s survival or censoring time. The Brier score provides a score that accounts for the information loss because of censoring. There are three categories of individuals that may appear within the dataset once we fix our $t$ of interest. Individuals who experienced the event before $t$ are present in the first term of the equation. The second term of the equation includes individuals who experience the event or are censored after $t$. Those censored before $t$ are the third category of people. The inverse probability censoring weights (IPCW) adjustment ($G(\cdot)$) is to account for these category three individuals whose information is missing. The IPA score as a function of time is given by
\begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where $BS_{model}(t)$ represents the Brier score over time $t$ for the model of interest and $BS_{null}(t)$ represents the Brier score if we use an unadjusted Kaplan-Meier (KM) curve as the prediction for all observations [@kattan2018index]. Note that IPA has an upper bound of one, where positive values show an increase in performance over the null model and negative values show that the null model performs better. These scores show how performance changes over follow-up time.

A potential artifact of IPA is that the score is unstable at earlier and later survival times. This is because of a near equivalent Brier score among each model and the null model. At small values, a difference of $0.1$ creates a more significant fold change than at larger values. As the Brier score is potentially small at the start and at the end of their respective curves, The IPA score may be unstable at the same locations.

@@

The $C_{IPCW}$ is a non-proper, rank-based metric that does not depend on the censoring times in the test data [@uno2011]. The $C_{IPCW}$ is given by
